|

### 2025-05-23 03-results.Rmd
edits

We tested whether showing students AI-generated step-by-step reasoning affected how they used the AI's prediction, compared to just seeing the prediction without any reasoning. For each student $i = 1, \cdots, 122$, we computed the absolute error as the absolute difference between student $i$'s prediction, $\text{Guess}_i$, and the actual value of coins:
$$
  \text{Absolute Error}_i = \mid \text{Guess}_i - \text{Actual Value Coins} \mid,
$$
where a smaller absolute error indicated greater accuracy.

@@

Figure \@ref(fig:boot-accuracy) illustrates that students' prediction accuracy was positively skewed in both treatment groups, with a greater accuracy for students exposed to AI reasoning. To estimate confidence intervals for the median difference in accuracy betweeen the groups, we used a nonparametric bootstrap approach.[^outlier] This procedure involved resampling participants' absolute errors (n = 1,999) to build a distribution of medians for each group. The differences between groups were computed, and the 5th and 95th percentiles of the resulting distribution were used to construct a 90% confidence interval for the difference in accuracy between groups. The interval ranged from 0 to 70, indicating a significantly greater accuracy (lower error) for the students exposed to AI-generated reasoning (one-sided, p < 0.05).

[^outlier]: We detected one outlier in the data, which was removed from the bootstrap analysis. One student provided a notably high guess of 6969. However, this value deviated substantially from both the overall distribution and the student's initial guess, suggesting a mistake rather than an estimate.

@@

Conversely, exposure to the AI step-by-step reasoning significantly influenced students' perceptions of the "human" estimate of $596. This value was the average guess of 600 participants in the initial study [@steiner2015turns], which exceeded the correct value ($379.54) by about the same amount as the AI guess underestimated it ($213). Notably, all groups received the same human and AI estimates, but while 64% of students in the control group rated the human estimate as "poor" or "very poor," only 45% of students exposed to AI reasoning did so. Although this difference was marginally significant (Fisher's exact test, p = 0.15),  regression analysis, controlling for student's location, gender, their initial estimate of the value of coins, and experience with ChatGPT, confirmed that AI exposure led students to rate the human estimate about half a point more favourably (p < 0.05), as shown in Figure \@ref(fig:perceived-accuracy-coins).  This finding underscores the complex relationship between AI and learning. It suggests that students exposed to the AI's step-by-step reasoning may have identified minor errors, thus increasing their expectations of human accuracy, and those who viewed AI as a "black box" tended to underrate human estimates.

>>>>

We tested whether showing students AI-generated step-by-step reasoning affected how they used the AI's prediction, compared to just seeing the prediction without any reasoning. For each student $i = 1, \cdots, 122$, we computed the absolute error as the absolute difference between student $i$'s guess, $\text{Guess}_i$, and the actual value of coins:
$$
  \text{Absolute Error}_i = \mid \text{Guess}_i - \text{Actual Value Coins} \mid,
$$
where a smaller absolute error indicated greater accuracy.

@@

Figure \@ref(fig:boot-accuracy) illustrates that students' prediction accuracy was positively skewed in both treatment groups, with a greater accuracy for students exposed to AI reasoning. To estimate confidence intervals for the median difference in accuracy betweeen the groups, we used a nonparametric bootstrap approach.[^outlier] This procedure involved resampling participants' absolute errors (n = 1,999) to build a distribution of medians for each group. The 5th and 95th percentiles of the resulting distribution were used to construct a 90% confidence interval for the difference in accuracy between groups. The interval ranged from 0 to 70, indicating a greater accuracy (lower error) for the students exposed to AI-generated reasoning (one-sided, p < 0.05).

[^outlier]: We detected one outlier in the data, which was removed from the bootstrap analysis: one student provided an unusual and notably high guess of 6969. However, this value deviated substantially from both the overall distribution and the student's initial guess, suggesting a mistake rather than an estimate.

@@

Conversely, exposure to the AI step-by-step reasoning seemed associated with students' perception of the accuracy of the "crowd" guess --- the average guess from 600 participants reported in the original study [@steiner2015turns].^[Notably, this guess was $596, exceeding the correct value ($379.54) by about the same amount as the AI guess underestimated it ($213).] All groups received the same human and AI estimates, but while 64% of students in the control group rated the human estimate as "poor" or "very poor," only 45% of students exposed to AI reasoning did so. Although this difference was insignificant (Fisher's exact test, p = 0.15),  regression analysis, controlling for student's location, gender, their initial estimate of the value of coins, and experience with ChatGPT, showed a significant effect (p < 0.05) of AI exposure on students' perceptions of the human estimate, as shown in Figure \@ref(fig:perceived-accuracy-coins). This finding underscores the complex relationship between AI and learning. It suggests that students exposed to the AI's step-by-step reasoning may have identified minor errors, thus increasing their expectations of human accuracy, and those who viewed AI as a "black box" tended to underrate human estimates.
